 

### 2023-03-08 markdown/Thesis.tex
Statistical testing in thesis started

---
title: " "
output: pdf_document
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
---
# Introduction

## Proteomics

@@

## Mass Spectrometry

@@

### Instrumentation

#### Cell Isolation

#### Lysis

#### Digestion

#### Liquid chromatography

@@

#### Ionization

In order to analyze a biological sample consisting of proteins in solution the liquid needs to be vaporized into gas phase. Two techniques are capable of this procedure. Electrospray ionization (ESI) pushes the analyte through a capillary and applies an electric current to the liquid, vaporizing the sample to a  charged aerosol. Biomolecules are fragmented according to their chemical properties and can be further handled in the mass spectrometer. The fragmented biomolecules are now in charged droplets separated by their charge on the surface, splitting further into smaller droplets until they become a gas phase ion. Two physical models describe the process from gas phase to ion called “The ion evaporation model” (IEM) and “The charge residue model” (CRM). 
In the ion evaporation model (by Iribarne and Thomoson) the droplets shrink by evaporation until ions are expelled. The model had its limitation by explaining same evaporation rate constant among ions with different chemical properties. 
In the charge residue model the assumption of one molecule per droplet leads to an ionization rate constant, which is independent of the ion itself and relies solely on the generation of the droplet and the efficiency of the solvent \citep{Wilm2011}.

@@

#### Coupled mass-spectrometry (MS/MS)

@@

#### Labeling techniques

@@

#### Metabolic labeling

@@

##### Isobaric labeling

###### Tandem mass tag (TMT)

@@

### MS-Spectrum

@@

### Interpretation of the data

#### Data acquisition

@@

##### Three-dimensional peak detection

@@

##### Deisotoping

@@

##### Label detection

The next step in data-acquisition is the detection of labels for quantification. Isotopic pairs of the label (e.g. N13, N14, N15) contained in the tag or aminoacid can be identified by convoluting the two measured isotope patterns with the theoretical isotope patterns. With a least-square method the best fit is found iteratively and the channel/sample can be identified.

##### Improving peptide mass accuracy

@@

##### Peptide search

@@

##### Protein assembly

@@

#### Data processing

@@

##### Reading the data

@@

* Peptide sequence
* Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)
	+ Mass
* Retention time
* Precursor Ion Fragment
	+ source of the detected ion also referred as mother ion
* Fraction of total spectrum 
* Base peak fraction
* Reporter intensity (RI)
	+ corrected RI
* Posterior error probability (PEP)

#### Object oriented programming

@@

#### Filtering

@@

The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides pectrum matching score is defined as -10log10(p). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda , an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

@@

#### Downstream Analysis

##### Principle component analysis (PCA)

@@

#### Testing for differential expression

According to the experiment and sample groups, individual statistical methods should be used.

follow sctranscr. analysis pipe

signatures of cell clustures

monocyte data acvailiabe

tmt label --> noise reduction

tools for clustering

	plst, knm,

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[left = 2.5cm, right = 2cm, top = 2cm, bottom =
2cm]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{float}
\usepackage{sectsty}
\usepackage{paralist}
\usepackage{setspace}\spacing{1.5}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{dcolumn}
\usepackage{natbib}\bibliographystyle{agsm}
\usepackage[nottoc, numbib]{tocbibind}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\allsectionsfont{\centering}
\subsectionfont{\raggedright}
\subsubsectionfont{\raggedright}

\pagenumbering{gobble}

\begin{centering}

\vspace{3cm}

\includegraphics[width=0.2\linewidth]{FHCW_logo} 
\includegraphics[width=0.2\linewidth]{KI_logo}

\vspace{1cm}

\Large
\doublespacing
{\bf Design and implementation of analysis pipeline for single cell type proteomics data}

\vspace{1 cm}

\normalsize
\singlespacing
By

\vspace{0.5 cm}

\Large

{\bf Lukas Gamp}

\vspace{1.5 cm}

in partial fulfilment of the requirement \\for the degree of MSc \\in Bioinformatics

\vspace{1.5 cm}

\normalsize
mm yy
\end{centering}

\newpage

\pagenumbering{gobble}

\begin{centering}

{\bf Abstract}

\end{centering}

\spacing{1.5}

(the spacing is set to 1.5)

no more than 250 words for the abstract

\begin{itemize}
\tightlist
\item
  a description of the research question/knowledge gap -- what we know
  and what we don't know
\item
  how your research has attempted to fill this gap
\item
  a brief description of the methods
\item
  brief results
\item
  key conclusions that put the research into a larger context
\end{itemize}

\pagenumbering{roman}

\newpage

\centering
\raggedright
\newpage
\tableofcontents

\newpage

\section*{Acknowledgements}

Thank you for following this tutorial!

I hope you'll find it useful to write a very professional dissertation.

\newpage

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{proteomics}{%
\subsection{Proteomics}\label{proteomics}}

@@

\hypertarget{mass-spectrometry}{%
\subsection{Mass Spectrometry}\label{mass-spectrometry}}

@@

\hypertarget{instrumentation}{%
\subsubsection{Instrumentation}\label{instrumentation}}

\hypertarget{cell-isolation}{%
\paragraph{Cell Isolation}\label{cell-isolation}}

\hypertarget{lysis}{%
\paragraph{Lysis}\label{lysis}}

\hypertarget{digestion}{%
\paragraph{Digestion}\label{digestion}}

\hypertarget{liquid-chromatography}{%
\paragraph{Liquid chromatography}\label{liquid-chromatography}}

@@

\hypertarget{ionization}{%
\paragraph{Ionization}\label{ionization}}

In order to analyze a biological sample consisting of proteins in
solution the liquid needs to be vaporized into gas phase. Two techniques
are capable of this procedure. Electrospray ionization (ESI) pushes the
analyte through a capillary and applies an electric current to the
liquid, vaporizing the sample to a charged aerosol. Biomolecules are
fragmented according to their chemical properties and can be further
handled in the mass spectrometer. The fragmented biomolecules are now in
charged droplets separated by their charge on the surface, splitting
further into smaller droplets until they become a gas phase ion. Two
physical models describe the process from gas phase to ion called ``The
ion evaporation model'' (IEM) and ``The charge residue model'' (CRM). In
the ion evaporation model (by Iribarne and Thomoson) the droplets shrink
by evaporation until ions are expelled. The model had its limitation by
explaining same evaporation rate constant among ions with different
chemical properties. In the charge residue model the assumption of one
molecule per droplet leads to an ionization rate constant, which is
independent of the ion itself and relies solely on the generation of the
droplet and the efficiency of the solvent \citep{Wilm2011}.

@@

\hypertarget{coupled-mass-spectrometry-msms}{%
\paragraph{Coupled mass-spectrometry
(MS/MS)}\label{coupled-mass-spectrometry-msms}}

@@

\hypertarget{labeling-techniques}{%
\paragraph{Labeling techniques}\label{labeling-techniques}}

@@

\hypertarget{metabolic-labeling}{%
\paragraph{Metabolic labeling}\label{metabolic-labeling}}

@@

\hypertarget{isobaric-labeling}{%
\subparagraph{Isobaric labeling}\label{isobaric-labeling}}

Tandem mass tag (TMT)

@@

\hypertarget{ms-spectrum}{%
\subsubsection{MS-Spectrum}\label{ms-spectrum}}

@@

\hypertarget{interpretation-of-the-data}{%
\subsubsection{Interpretation of the
data}\label{interpretation-of-the-data}}

\hypertarget{data-acquisition}{%
\paragraph{Data acquisition}\label{data-acquisition}}

@@

\hypertarget{three-dimensional-peak-detection}{%
\subparagraph{Three-dimensional peak
detection}\label{three-dimensional-peak-detection}}

@@

\hypertarget{deisotoping}{%
\subparagraph{Deisotoping}\label{deisotoping}}

@@

\hypertarget{label-detection}{%
\subparagraph{Label detection}\label{label-detection}}

The next step in data-acquisition is the detection of labels for
quantification. Isotopic pairs of the label (e.g.~N13, N14, N15)
contained in the tag or aminoacid can be identified by convoluting the
two measured isotope patterns with the theoretical isotope patterns.
With a least-square method the best fit is found iteratively and the
channel/sample can be identified.

\hypertarget{improving-peptide-mass-accuracy}{%
\subparagraph{Improving peptide mass
accuracy}\label{improving-peptide-mass-accuracy}}

@@

\hypertarget{peptide-search}{%
\subparagraph{Peptide search}\label{peptide-search}}

@@

\hypertarget{protein-assembly}{%
\subparagraph{Protein assembly}\label{protein-assembly}}

@@

\hypertarget{data-processing}{%
\paragraph{Data processing}\label{data-processing}}

@@

\hypertarget{reading-the-data}{%
\subparagraph{Reading the data}\label{reading-the-data}}

@@

\begin{itemize}
\tightlist
\item
  Peptide sequence
\item
  Mass to charge ratio (m/z) for all scans (eg. MS1, MS2)

  \begin{itemize}
  \tightlist
  \item
    Mass
  \end{itemize}
\item
  Retention time
\item
  Precursor Ion Fragment

  \begin{itemize}
  \tightlist
  \item
    source of the detected ion also referred as mother ion
  \end{itemize}
\item
  Fraction of total spectrum
\item
  Base peak fraction
\item
  Reporter intensity (RI)

  \begin{itemize}
  \tightlist
  \item
    corrected RI
  \end{itemize}
\item
  Posterior error probability (PEP)
\end{itemize}

\hypertarget{object-oriented-programming}{%
\paragraph{Object oriented
programming}\label{object-oriented-programming}}

@@

\hypertarget{filtering}{%
\paragraph{Filtering}\label{filtering}}

@@

The next step for quality control is the exclusion of samples with a
high false discovery rate (FDR). When applying multiple statistical
testing (e.g.~t-Test) the obtained p-values can be considered as biased,
because the probability to observe a significant will iteratively
increase with each test performed. Corrections in statistics are an
approach to compensate for the multiplicity of testing. There are many
ways to do this compensation like the Bonferroni method or
Benjamini-Hochberg`s FDR. In Mass spectrometry the common ``way to go''
is calculating a false discovery rate, by dividing false PSMs (=hit of
the decoy database) through the total number of PSMs above the
peptide-spectrum matching score. The peptides spectrum matching score is
defined as -10log10(p). Whereas the p-value is defined that the hit is
done by chance. The calculation of the score is highly dependent on the
data acquisition method used. MaxQuant uses Andromeda , an integrated
search engine. Proteome Discoverer from Thermo Fisher utilizes different
engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot
and Andromeda showed similar performance when comparing FDR values as a
function of coverage. However the observed performance can be lower when
dealing with a decreased coverage \citep{Cox2011}.The threshold for
accepting an FDR of an individual PSM is described as q-value.

@@

\hypertarget{downstream-analysis}{%
\paragraph{Downstream Analysis}\label{downstream-analysis}}

\hypertarget{principle-component-analysis-pca}{%
\subparagraph{Principle component analysis
(PCA)}\label{principle-component-analysis-pca}}

@@

\hypertarget{testing-for-differential-expression}{%
\paragraph{Testing for differential
expression}\label{testing-for-differential-expression}}

In order to test for differential expression the package limma from R
bioconductor was used. \citep{Phipson2016}. The package uses a linear
model approach to define a fold expression between sample groups. Before
starting the analysis two matrices need to be computed. The design
matrix identifies samples according to the sample type and defines the
experimental design. In order to create the matrix automatically a
simple algorithm is used within the programmed backend logic. The
expression matrix contains intensities for each identified protein will
be obtained at the end of the pipeline. A function call with the
arguments design and expression matrix calculates the contrast matrix.
The next step is creating a linear model between groups by log-ratios of
their expression values.

\pagenumbering{arabic}

\newpage

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\vspace{0.5cm}

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

\hypertarget{materials}{%
\subsection{Materials}\label{materials}}

For analysis two types of cells were used. One type is the Jurkat-based
cell line (J-lat) with integrated HIV.

The other type of cells are macrophages with a sample size of 72 cells.
The analysis is done with two groups. A HIV negative (HIV-) control
group and a HIV positive (HIV+) group.

\hypertarget{data}{%
\subsection{Data}\label{data}}

\hypertarget{acquisition}{%
\subsubsection{Acquisition}\label{acquisition}}

Analysis of the data was done with MaxQuant \citep{Cox2008}

\newpage

\hypertarget{results}{%
\section{Results}\label{results}}

Some more guidlines from the School of Geosciences.

This section should summarise the findings of the research referring to
all figures, tables and statistical results (some of which may be placed
in appendices). - include the primary results, ordered logically - it is
often useful to follow the same order as presented in the methods. -
alternatively, you may find that ordering the results from the most
important to the least important works better for your project. - data
should only be presented in the main text once, either in tables or
figures; if presented in figures, data can be tabulated in appendices
and referred to at the appropriate point in the main text.

\textbf{Often, it is recommended that you write the results section
first, so that you can write the methods that are appropriate to
describe the results presented. Then you can write the discussion next,
then the introduction which includes the relevant literature for the
scientific story that you are telling and finally the conclusions and
abstract -- this approach is called writing backwards.}

\newpage

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

the purpose of the discussion is to summarise your major findings and
place them in the context of the current state of knowledge in the
literature. When you discuss your own work and that of others, back up
your statements with evidence and citations. - The first part of the
discussion should contain a summary of your major findings (usually 2 --
4 points) and a brief summary of the implications of your findings.
Ideally, it should make reference to whether you found support for your
hypotheses or answered your questions that were placed at the end of the
introduction. - The following paragraphs will then usually describe each
of these findings in greater detail, making reference to previous
studies. - Often the discussion will include one or a few paragraphs
describing the limitations of your study and the potential for future
research. - Subheadings within the discussion can be useful for
orienting the reader to the major themes that are addressed.

\newpage

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

The conclusion section should specify the key findings of your study,
explain their wider significance in the context of the research field
and explain how you have filled the knowledge gap that you have
identified in the introduction. This is your chance to present to your
reader the major take-home messages of your dissertation research. It
should be similar in content to the last sentence of your summary
abstract. It should not be a repetition of the first paragraph of the
discussion. They can be distinguished in their connection to broader
issues. The first paragraph of the discussion will tend to focus on the
direct scientific implications of your work (i.e.~basic science,
fundamental knowledge) while the conclusion will tend to focus more on
the implications of the results for society, conservation, etc.

\newpage

\bibliography{bibliography}

\newpage

\hypertarget{appendixces}{%
\section{Appendix(ces)}\label{appendixces}}

\hypertarget{appendix-a-additional-tables}{%
\subsection{Appendix A: additional
tables}\label{appendix-a-additional-tables}}

Insert content for additional tables here.

\newpage

\hypertarget{appendix-b-additional-figures}{%
\subsection{Appendix B: additional
figures}\label{appendix-b-additional-figures}}

Insert content for additional figures here.

\newpage

\hypertarget{appendix-c-code}{%
\subsection{Appendix C: code}\label{appendix-c-code}}

Insert code (if any) used during your dissertation work here.

\end{document}

### 2023-03-08 markdown/introduction.Rmd
Statistical testing in thesis started

The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides pectrum matching score is defined as -10log10(p). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda , an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

@@

According to the experiment and sample groups, individual statistical methods should be used.

follow sctranscr. analysis pipe

signatures of cell clustures

monocyte data acvailiabe

tmt label --> noise reduction

tools for clustering

	plst, knm,

>>>>

The next step for quality control is the exclusion of samples with a high false discovery rate (FDR). When applying multiple statistical testing (e.g. t-Test) the obtained p-values can be considered as biased, because the probability to observe a significant will iteratively increase with each test performed. Corrections in statistics are an approach to compensate for the multiplicity of testing. There are many ways to do this compensation like the Bonferroni method or Benjamini-Hochberg`s FDR. In Mass spectrometry the common “way to go” is calculating a false discovery rate, by dividing false PSMs (=hit of the decoy database) through the total number of PSMs above the peptide-spectrum matching score. The peptides spectrum matching score is defined as -10log10(p). Whereas the p-value is defined that the hit is done by chance. The calculation of the score is highly dependent on the data acquisition method used. MaxQuant uses Andromeda , an integrated search engine. Proteome Discoverer from Thermo Fisher utilizes different engines such as Mascot or Minora. As published by J.Cox in 2011 Mascot and Andromeda showed similar performance when comparing FDR values as a function of coverage. However the observed performance can be lower when dealing with a decreased coverage \citep{Cox2011}.The threshold for accepting an FDR of an individual PSM is described as q-value.

@@

In order to test for differential expression the package limma from R bioconductor was used. \citep{Phipson2016}.  The package uses a linear model approach to define a fold expression between sample groups. Before starting the analysis two matrices need to be computed. The design matrix identifies samples according to the sample type and defines the experimental design. In order to create the matrix automatically a simple algorithm is used within the programmed backend logic. The expression matrix contains intensities for each identified protein will be obtained at the end of the pipeline. A function call with the arguments design and expression matrix calculates the contrast matrix. The next step is creating a linear model between groups by log-ratios of their expression values.
